---
title: ''
output: html_document
---

# 📈 **Linear Models**

## **0.** Settings
```{r, warning=FALSE, message=FALSE}
library(MASS)
library(car)
library(rgl)
```

## **Part 1 - Linear Models, Plots, Clustering**

## **1.** Dataset
```{r, warning=FALSE, message=FALSE}
data = cars
data[, 'speed^2'] = cars$speed^2
data = data[,-2]
data[, 'dist'] = cars$dist
head(data)

# Re-name
n = dim(data)[1]
p = dim(data)[2]-1
v = 'X1'
if(p>1){
  for(i in 2:p){
    v = c(v, paste('X',i,sep=''))
  }
}
v = c(v, 'Y')
v
colnames(data) = v
head(data)
attach(data)
```

## **2.** Exploring (p=2)
```{r, warning=FALSE, message=FALSE}
plot(X1, Y, xlab='X1', ylab='Y')
plot(X2, Y, xlab='X2', ylab='Y')
pairs(data)
#plot3d(X1, X2, Y, size=4, asp = T)
```

## **3.** Model
```{r, warning=FALSE, message=FALSE}
# Y = beta_0 + beta_1 * X1 + beta_2 * X2 + Eps

## Assumptions:
# 1) Parameter estimation: E(Eps) = 0 and Var(Eps) = sigma^2
# 2) Inference : Eps ~ N(0, sigma^2)

# Estimate the parameters
fm = lm(Y ~ X1 + X2)
summary(fm)
vif(fm)

# Comment
# * Residual standard error = estimate of sigma
# * Degrees of freedom = n-(r+1)
# * F-statistic - p.value = H0: beta_1 = .. = beta_r = 0
#   (low p-value -> it makes sense to go on)
# * Check for R^2 and R.adj^2
# * Residuals have to be symmetric 
# * Pr(>|t|) = p-value of the one-at-time beta test
#   !ATTENTION! Delete one-at-the-time

# Y hat
fitted(fm)

# Eps hat
residuals(fm)
plot(residuals(fm))

# Beta hat
coefficients(fm)

# Cov(beta hat) = sigma^2 * (Z^T * Z)^(-1)
vcov(fm)

# Order of the model (r+1)
fm$rank

# Degrees of freedom for the residuals
fm$df

# Leverages h_ii
hatvalues(fm)

# Standardized residuals
rstandard(fm)
plot(rstandard(fm))

# Estimate of sigma^2
sum(residuals(fm)^2)/fm$df
```

## **4.** Regression plot
```{r, warning=FALSE, message=FALSE}
# (specific case of a regression of the form: Y = b_0 + b_1 * X + b_2 * X^2)

x = seq(0,30,by=0.01)
b = coef(fm)
plot(X1, Y, xlab='X1', ylab='Y')
lines(x, b[1]+b[2]*x+b[3]*x^2)

# If p=2: 3D plot
#par3d(windowRect=c(680,40,1350,720))
#points3d(x=X1, y=X2, z=Y, size=4, aspect = T)
#box3d()
#axes3d()
#points3d(x=X1, y=X2, z=fitted(fm), size=4, col = 'blue')
#surface3d(range(data$X1), range(data$X2),
#          matrix(predict(fm, expand.grid(X1=range(X1), X2=range(X2))),2,2),alpha = 0.5)
```

## **5.** Inference on Betas
```{r, warning=FALSE, message=FALSE}
## Assuption: Eps ~ N(0, sigma^2)
## Test (Fisher):
##    H0: (beta1, beta2) == (0, 0)
##    H1: (beta1, beta2) != (0, 0)

linearHypothesis(fm, rbind(c(0,1,0), c(0,0,1)), c(0,0))

# With more betas (e.g. 3):
# linearHypothesis(fm, rbind(c(0,1,0,0), c(0,0,1,0), c(0,0,0,1)), c(0,0,0))

# Comment
#   Pr(>F) = final p-value in summary(fm)

r = fm$rank - 1
```

## **6.** Confidence Regions (if p=2)
```{r, warning=FALSE, message=FALSE}
# Center
c(coefficients(fm)[2], coefficients(fm)[3])

# Direction of the axes
eigen(vcov(fm)[2:3, 2:3])$vectors
plot(coefficients(fm)[2], coefficients(fm)[3], xlim = c(-6,6), ylim = c(-6,6),
     asp=1, xlab='beta1', ylab='beta2')
ellipse(coefficients(fm)[2:3], vcov(fm)[2:3,2:3], sqrt(p*qf(1-0.05,p,n-(r+1))))
abline(v=0)
abline(h=0)

# Comment
#   If it is 'squished' then (probably) they're collinear.
```

## **7.** Bonferroni Intervals (level 95%, p=2)
```{r, warning=FALSE, message=FALSE}
Bf = rbind(
  beta1=c(coefficients(fm)[2]-sqrt(vcov(fm)[2,2])*qt(1-0.05/(2*p), n-(r+1)),
          coefficients(fm)[2]+sqrt(vcov(fm)[2,2])*qt(1-0.05/(2*p), n-(r+1))),
  beta2=c(coefficients(fm)[3]-sqrt(vcov(fm)[3,3])*qt(1-0.05/(2*p), n-(r+1)),
          coefficients(fm)[3]+sqrt(vcov(fm)[3,3])*qt(1-0.05/(2*p), n-(r+1))))
Bf

# Generic beta_j (p, r, n, alpha generici)
# beta_j = c(coefficients(fm)[j]-sqrt(vcov(fm)[j,j])*qt(1-alpha/(2*p), n-(r+1)),
# coefficients(fm)[j]+sqrt(vcov(fm)[j,j])*qt(1-alpha/(2*p), n-(r+1)))

# Alternatively: Bonferroni's correction
confint(fm, level= 1-0.05/p)[2:3,]
```

## **8.** Other tests with "linearHypothesis"
```{r, warning=FALSE, message=FALSE}
# Test:
#   H0: (beta0+beta2, beta1) == (0,0)
#   H1: (beta0+beta2, beta1) != (0,0)

C = rbind(c(1,0,1), c(0,1,0))
linearHypothesis(fm, C, c(0,0))

# Simultaneous CI (beta_j combinations):

# Fisher 1-alpha quantile 
alpha = 0.05
qf.fish = qf(1-alpha, r+1, n-(r+1))

# Customized combination (here we just consider beta_1)
a = c(0,1,0)

# Confidence Interval
sim_IC = c(t(a)%*%coefficients(fm) - sqrt(t(a)%*%vcov(fm)%*%a) * sqrt((r+1)*qf.fish),
           t(a)%*%coefficients(fm) + sqrt(t(a)%*%vcov(fm)%*%a) * sqrt((r+1)*qf.fish))
sim_IC
```

## **9.** Inference on the mean (CI(E[Y|X]), PI(Y))
```{r, warning=FALSE, message=FALSE}
# New sample
Z0.new = data.frame(X1=10, X2=10^2)

alpha = 0.05

# CI(E[Y|X])
Conf = predict(fm, Z0.new, interval='confidence', level=1-alpha)
Conf

# PI(Y)
Pred = predict(fm, Z0.new, interval='prediction', level=1-alpha)
Pred
```

## **10.** Plot IC vs IP: one sample
```{r, warning=FALSE, message=FALSE}
# Particularity for this model case
plot(X1, Y, xlab='X1', ylab='Y', las=1)
x = seq(0,30,by=0.1)
b = coef(fm)
lines(x, b[1]+b[2]*x+b[3]*x^2)
points(10,Conf[1], pch=19)
segments(10,Pred[2],10,Pred[3],col='gold', lwd=2)
segments(10,Conf[2],10,Conf[3],col='red', lwd=2)
points(10,Conf[2], pch='-', col='red', lwd=2)
points(10,Conf[3], pch='-', col='red', lwd=2)
points(10,Pred[2], pch='-', col='gold', lwd=2)
points(10,Pred[3], pch='-', col='gold', lwd=2)
```

## **11.** Plot IC vs IP: grid of sample
```{r, warning=FALSE, message=FALSE}
Z0 = data.frame(cbind(X1 = seq(min(X1), max(X1), length=100),
                      X2 = seq(min(X2), max(X2), length=100)))
Conf = predict(fm, Z0, interval='confidence')
Pred = predict(fm, Z0, interval='prediction')
plot(X1, Y, xlab='X1', ylab='Y', las=1)
lines(Z0[,1], Conf[,'fit'])
lines(Z0[,1], Conf[,'lwr'], lty=2, col='red', lwd=2)
lines(Z0[,1], Conf[,'upr'], lty=2, col='red', lwd=2)
lines(Z0[,1], Pred[,'lwr'], lty=3, col='gold', lwd=2)
lines(Z0[,1], Pred[,'upr'], lty=3, col='gold', lwd=2)

# Comment
#   These are NOT confidence/prediction BANDS, they're done one-at-the-time.
```

## **12.** Verify assumptions (used for inference and estimate of parameters)
```{r, warning=FALSE, message=FALSE}
# * Gaussianity
# * Homoschedasticity

par(mfrow=c(2,2))
plot(fm)

# Comment
#   1. We want to see no pattern: a cloud around the zero
#   2. We want to see a good fit on the line
#   3. Again, we want to see no pattern
#   4. We have the iso-lines of the Cook distance: we can identify the outliers

shapiro.test(residuals(fm))
detach(data)
```

## **13.** Data transformation (another example)
```{r, warning=FALSE, message=FALSE}
data = read.table('brain_weight.txt', head=T)
head(data)

# Re-name
n = dim(data)[1]
p = dim(data)[2]-1
v = 'X1'
if(p>1){
  for(i in 2:p){
    v = c(v, paste('X',i,sep=''))
  }
}
v = c(v, 'Y')
v
colnames(data) = v
head(data)
attach(data)

# Linear regression
logX1 = log(X1)
logY  = log(Y)
fm    = lm(logY~logX1)

# Plot
plot(logX1, logY)
abline(coef(fm)[1], coef(fm)[2])

# Plot IC vs IP: transformed datas' grid
logZ0 = data.frame(logX1=seq(min(logX1), max(logX1), length=100))
Conf  = predict(fm, logZ0, interval='confidence')
Pred  = predict(fm, logZ0, interval='prediction')

plot(logX1, logY, xlab='logX1', ylab='logY', las=1)
lines(logZ0[,1], Conf[,'fit'])
lines(logZ0[,1], Conf[,'lwr'], lty=2, col='red', lwd=2)
lines(logZ0[,1], Conf[,'upr'], lty=2, col='red', lwd=2)
lines(logZ0[,1], Pred[,'lwr'], lty=3, col='gold', lwd=2)
lines(logZ0[,1], Pred[,'upr'], lty=3, col='gold', lwd=2)

# Comment
#   These are NOT confidence/prediction BANDS, they're done one-at-the-time.

# Plot IC vs IP: transformed datas' grid
plot(X1, Y, xlab='X1', ylab='Y', las=1)
CI = exp(Conf)
PI = exp(Pred)
Z0 = exp(logZ0)

lines(Z0[,1], CI[,'fit'])
lines(Z0[,1], CI[,'lwr'], lty=2, col='red', lwd=2)
lines(Z0[,1], CI[,'upr'], lty=2, col='red', lwd=2)
lines(Z0[,1], PI[,'lwr'], lty=3, col='gold', lwd=2)
lines(Z0[,1], PI[,'upr'], lty=3, col='gold', lwd=2)

# Comment
#   These are NOT confidence/prediction BANDS, they're done one-at-the-time.

detach(data)
```

## **14.** Desperate Case: Introduction of Dummy Variables
```{r, warning=FALSE, message=FALSE}
# If the model is terrible, we can try to perform a hierarchical clustering and 
# add some dummy vairbales (given by the cluster)

data = read.table('brain_weight.txt', head=T)
head(data)

# Re-name
n = dim(data)[1]
p = dim(data)[2]-1
v = 'X1'
if(p>1){
  for(i in 2:p){
    v = c(v, paste('X',i,sep=''))
  }
}
v = c(v, 'Y')
v
colnames(data) = v
head(data)

data['X2'] = data['X1']^2
head(data)

# Cluster method: ward
clusterw = cutree(hclust(dist(data), method='ward.D2'),2)
dummy = clusterw - 1

# Model
# Y = b_0 + b_1*X1 + b_2*X2 + b_3*dummy + b_4*dummy*X1 + b_5*dummy*X2
fm2 = lm(Y ~ X1 + X2 + dummy + X1:dummy + X2:dummy, data=data)
summary(fm2)

# Plot
par(mfrow=c(2,2))
plot(fm2)
```

## **Part 2 - Collinearity:,PCA, Ridge, Lasso**

## **1.** Dataset
```{r, warning=FALSE, message=FALSE}
data = cars
data[, 'speed^2'] = cars$speed^2
data = data[,-2]
data[, 'dist'] = cars$dist
head(data)

# Re-name
n = dim(data)[1]
p = dim(data)[2]-1
v = 'X1'
if(p>1){
  for(i in 2:p){
    v = c(v, paste('X',i,sep=''))
  }
}
v = c(v, 'Y')
v
colnames(data) = v
head(data)
attach(data)
```

## **2.** Model
```{r, warning=FALSE, message=FALSE}
fm = lm(Y ~ X1 + X2)
summary(fm)

# Check for collinearity
vif(fm)

# Note
#   If it is >10 then it is considered 'high'
```

## **3.** PCA Regression (p=2)
```{r, warning=FALSE, message=FALSE}
# PCA
data.pc = princomp(cbind(X1,X2), scores=TRUE)
summary(data.pc)

# Note
#   Based on 'Cumulative Proportion' we decide how many components to consider.

# PCA regression
#   Y = a_0 + a_1 * X1.pc + a_2 * X2.pc
X1.pc = data.pc$scores[,1]
X2.pc = data.pc$scores[,2]
fm.pc = lm(Y ~ X1.pc + X2.pc)
summary(fm.pc)

# Back to original variables
#   Y = b_0 + b_1 * X1 + b_2 * X2
# Transforming PCA regression:
#   b_0 = a_0
#   b_j = (e_j1 * a_1 + .. + e_jp * a_p )
# we create an alphas vector: [a_1, a_2, .., a_k]
k = length(fm.pc$coefficients)-1
alphas = as.vector(fm.pc$coefficients)
alpha0 = alphas[1]
alphas = alphas[2:(k+1)]
alphas = as.matrix(alphas) # (k x 1)

# and we take the loadings that are needed (first k)
E = as.matrix(data.pc$loadings[,])
E = as.matrix(E[,1:k]) # (p x k)

# we create the betas vector: [b_1, .., b_p]
betas = as.matrix(0*(1:p)) # (p x 1)

# we create the vetor of the means: [m1, .., mp]
m = 0*(1:p)
m[1] = mean(X1)
m[2] = mean(X2)
m = as.matrix(m) # (p x 1)

# beta_1, .., beta_p
betas = E %*% alphas

# beta 0
b_0 = alpha0 - t(m) %*% betas

# beta together
# b_0, b_1, .., b_p
beta_finali = c(b_0, betas)
beta_finali
```

## **4.** Ridge Regression
```{r, warning=FALSE, message=FALSE}
lambda = 0.5
fm.ridge = lm.ridge(Y ~ X1 + X2, lambda = lambda)

# Cofficients
coef(fm.ridge)

# Y.hat
y.hat.ridge = cbind(rep(1,n), X1, X2) %*% coef(fm.ridge)

# Optimal lambda
lambda.c=seq(0,10,0.01)
fm.ridge = lm.ridge(Y ~ X1 + X2, lambda = lambda.c)
select(fm.ridge)

# Alternatives (optimal lambda)
lambda.c=seq(0,10,0.01)
fm.ridge = lm.ridge(Y ~ X1 + X2, lambda = lambda.c)
lambda.opt = lambda.c[which.min(fm.ridge$GCV)]

# Cofficients for optimal lambda
coef.ridge = coef(fm.ridge)[which.min(fm.ridge$GCV),]
coef.ridge
```

## **5.** Ridge Regression (2)
```{r, warning=FALSE, message=FALSE}
library(glmnet)
x = model.matrix(Y ~ X1 + X2)[,-1] # matrix of predictors
y = Y # vector of response
lambda.grid = 10^seq(5,-3,length=50)

# Ridge regression
fit.ridge = glmnet(x,y, lambda = lambda.grid, alpha=0) # alpha=0 -> ridge
plot(fit.ridge,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
legend('topright', dimnames(x)[[2]], col = rainbow(dim(x)[2]), lty=1, cex=1)

# Set lambda via CV
cv.ridge = cv.glmnet(x,y,alpha=0,nfolds=3,lambda=lambda.grid)
bestlam.ridge = cv.ridge$lambda.min
bestlam.ridge

plot(cv.ridge)
abline(v=log(bestlam.ridge), lty=1)
coef.ridge <- predict(fit.ridge, s=bestlam.ridge, type = 'coefficients')[1:(p+1),]
coef.ridge
plot(fit.ridge,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
abline(v=log(bestlam.ridge))
```

## **6.** Lasso Regression
```{r, warning=FALSE, message=FALSE}
library(glmnet)
x = model.matrix(Y ~ X1 + X2)[,-1] # matrix of predictors
y = Y # vector of response
lambda.grid = 10^seq(5,-3,length=50)

# Lasso regression
fit.lasso = glmnet(x,y, lambda = lambda.grid, alpha=1) # alpha=1 -> lasso
plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
legend('topright', dimnames(x)[[2]], col = rainbow(dim(x)[2]), lty=1, cex=1)

# Set lambda via CV
cv.lasso = cv.glmnet(x,y,alpha=1,nfolds=3,lambda=lambda.grid)
bestlam.lasso = cv.lasso$lambda.min
bestlam.lasso

plot(cv.lasso)
abline(v=log(bestlam.lasso), lty=1)

# Get the coefficients for the optimal lambda
coef.lasso = predict(fit.lasso, s=bestlam.lasso, type = 'coefficients')[1:(p+1),]
coef.lasso

plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
abline(v=log(bestlam.lasso))
```

## **7.** Compare coefficients estimates for LS, Ridge and Lasso
```{r, warning=FALSE, message=FALSE}
plot(rep(0, dim(x)[2]), coef(fm)[-1], col=rainbow(dim(x)[2]), pch=20,
     xlim=c(-1,3), ylim=c(-1,2), xlab='', ylab=expression(beta),
     axes=F)
points(rep(1, dim(x)[2]), coef.ridge[-1], col=rainbow(dim(x)[2]), pch=20)
points(rep(2, dim(x)[2]), coef.lasso[-1], col=rainbow(dim(x)[2]), pch=20)
abline(h=0, col='grey41', lty=1)
box()
axis(2)
axis(1, at=c(0,1,2), labels = c('LS', 'Ridge', 'Lasso'))
legend('topright', dimnames(x)[[2]], col = rainbow(dim(x)[2]), pch=20, cex=1)
```

