---
title: ''
output: html_document
---

# 📈 **Supervised learning: LDA (uni/bivariate), QDA (biv), KNN, Fisher's argument**

## **0.** Settings
```{r, warning=FALSE, message=FALSE}
load('mcshapiro.test.RData')
library(MASS)
library(class)
library(rgl)
library(mvtnorm)
```

## **1.** LDA, KNN (2 classes, univariate)
```{r setup, include=FALSE}
data = read.table('cytokines.txt', header=T)

# Let's consider just the firs column and the label
data = data[,c(1,3)]
colnames(data) = c('X1', 'label')
data$label = as.factor(data$label)
head(data)
```

### **1.1.** Explore
```{r, warning=FALSE, message=FALSE}
livelli = levels(as.factor(data$label))
A = which(data$label == livelli[1])
B = which(data$label == livelli[2])
g = 2
p = 2
```

### **1.2.** LDA (univariate)
```{r, warning=FALSE, message=FALSE}
# Assumptions:
# 1) if L=i|X.i ~ N(mu.i, sigma.i^2), i=A,B
# 2) sigma.A = sigma.B
# 3) c(A|B)=c(B|A) (equal misclassification costs)

# 1)
shapiro.test(data[A,1])
shapiro.test(data[B,1])

# 2) (H0: same variance)
var.test(data[A,1], data[B,1])
var.test(data[A,1], data[B,1])$p.value
nA = length(A)
nB = length(B)
n  = nA + nB

# Prior probabilities
PA = nA/n
PB = nB/n

# LDA
data.lda = lda(label ~ X1, data)
data.lda
```

### **1.3.** Posterior probability and classification for x=0
```{r, warning=FALSE, message=FALSE}
is(data)
x = data.frame(X1 = 0)

# prediction
predict(data.lda, x)$class

# posterior probabilities for the class
predict(data.lda, x)$posterior

# Fisher's discriminant score
predict(data.lda, x)$x
```

### **1.4.** Graphical representation
```{r, warning=FALSE, message=FALSE}
# Training + posterior probabilities for x
plot(data$X1[A], rep(0, length(A)), pch=16, col='blue', ylim=c(0,1),
     xlab='x', ylab='estimated posterior', main="LDA", xlim=range(data$X1))
points(data$X1[B], rep(0, length(B)), pch=16, col='red')
abline(v=0, col='grey')
points(c(0,0),c(predict(data.lda, data.frame(X1 = 0))$posterior),
       col=c('blue', 'red'), pch='*', cex=2.5)

# Estimate the posterior probabilities for a grid
# (grid of representation)
x = data.frame(X1=seq(-10, 35, 0.5))
data.LDA.A <- predict(data.lda, x)$posterior[,1] # posterior probability for class A
data.LDA.B <- predict(data.lda, x)$posterior[,2] # posterior probability for class B

# Add to the plot
lines(x[,1], data.LDA.A, type='l', col='blue', xlab='x',
      ylab='estimated posterior', main="LDA")
lines(x[,1], data.LDA.B, type='l', col='red')
abline(h = 0.5)
legend(-10, 0.9, legend=c('P(A|X=x)', 'P(B|X=x)'), fill=c('blue','red'), cex = 0.7)
```

### **1.5.** Different prior probabilities?
```{r, warning=FALSE, message=FALSE}
data.lda.1 <- lda(label ~ X1, data, prior=c(0.95,0.05))
data.lda.1

## What happens in graphical representation?
# (the grey are the old probabilities (with priors not fixed by us))
x = data.frame(X1=seq(-10, 35, 0.5))
data.LDA.A.1 = predict(data.lda.1, x)$posterior[,1] # posterior probability for class A
data.LDA.B.1 = predict(data.lda.1, x)$posterior[,2] # posterior probability for class B
plot (x[,1], data.LDA.A.1, type='l', col='blue', xlab='x',
      ylab='estimated posterior', main="LDA", ylim=c(0,1))
points(x[,1], data.LDA.B.1, type='l', col='red')
abline(h = 0.5)
legend(-10, 0.9, legend=c('P(A|X=x)', 'P(B|X=x)'), fill=c('blue','red'), cex = 0.7)
points(data$X1[A], rep(0, length(A)), pch=16, col='blue')
points(data$X1[B], rep(0, length(B)), pch=16, col='red')
points(x[,1], data.LDA.A, type='l', col='grey')
points(x[,1], data.LDA.B, type='l', col='grey')
```

### **1.6.** Univariate LDA vs. KNN
```{r, warning=FALSE, message=FALSE}
data.knn = knn(train = data$X1, test = x, cl = data$label, k = 3, prob=T)
data.knn.class = (data.knn == 'B')+0
data.knn.B = ifelse(data.knn.class==1,
                    attributes(data.knn)$prob,
                    1 - attributes(data.knn)$prob)
plot(x[,1], data.LDA.B, type='l', col='red', lty=2, xlab='x', ylab='estimated posterior')
points(x[,1], data.knn.B, type='l', col='black', lty=1)
abline(h = 0.5)
legend(-10, 0.75, legend=c('LDA','knn'), lty=c(2,1), col=c('red','black'))

# Change k
par(mfrow=c(3,4))
for(k in 1:12){
  data.knn = knn(train = data$X1, test = x, cl = data$label, k = k, prob=T)
  data.knn.class = (data.knn == 'B')+0
  data.knn.B <- ifelse(data.knn.class==1,
                       attributes(data.knn)$prob,
                       1 - attributes(data.knn)$prob)
  plot(x[,1], data.LDA.B, type='l', col='red', lty=2, xlab='x',
       ylab='estimated posterior', main=k)
  points(x[,1], data.knn.B, type='l', col='black', lty=1, lwd=2)
  abline(h = 0.5)
}
```

## **2.** LDA, QDA, KNN (3 classes, bivariate: p = 2, g = 3)
```{r, warning=FALSE, message=FALSE}
data = iris[,c(1,2,5)]
colnames(data) = c('X1', 'X2', 'label')
head(data)
attach(data)

labels = as.factor(data$label)
tipi   = levels(data$label)
g      = length(tipi)

i1 = which(label==tipi[1])
i2 = which(label==tipi[2])
i3 = which(label==tipi[3])
n1 = length(i1)
n2 = length(i2)
n3 = length(i3)
n  = n1+n2+n3
detach(data)

plot(data[,1:2], main='Title', xlab='X1', ylab='Sepal.Width', pch=19)
points(data[i1,], col='red', pch=19)
points(data[i2,], col='green', pch=19)
points(data[i3,], col='blue', pch=19)
legend("topright", legend=levels(data$label), fill=c('red','green','blue'))

data = data[,1:2]
m  = colMeans(data)
m1 = colMeans(data[i1,])
m2 = colMeans(data[i2,])
m3 = colMeans(data[i3,])
S1 = cov(data[i1,])
S2 = cov(data[i2,])
S3 = cov(data[i3,])
Sp = ((n1-1)*S1+(n2-1)*S2+(n3-1)*S3)/(n-g)

# Check if the labels are effective on the variables: 
# One-way MANOVA
fit = manova(as.matrix(data) ~ labels)
summary.manova(fit, test="Wilks")
```

### **2.1.** LDA (multivariate)
```{r, warning=FALSE, message=FALSE}
lda.data = lda(data, labels)
lda.data

# "coefficients of linear discriminants" and "proportion of trace":
# Fisher discriminant analysis.
# In particular:
# - coefficients of linear discriminants: versors of the canonical directions
# [to be read column-wise]
# - proportion of trace: proportion of variance explained by the corresponding
# canonical direction
```

### **2.2.** Estimate of AER
```{r, warning=FALSE, message=FALSE}
# 1. APER
# 2. AER by cross-validation
#
# 1)
Lda.data = predict(lda.data, data)
names(Lda.data)
table(class.true=labels, class.assigned=Lda.data$class)
errors = (Lda.data$class != labels)
sum(errors)
length(labels)
APER = sum(errors)/length(labels)
APER

#### NOTE: if there are priors given:
# prior = c(1/3,1/3,1/3)
# G = 3
# misc = table(class.true=labels, class.assigned=Lda.data$class)
# APER = 0
# for(g in 1:G)
# APER = APER + sum(misc[g,-g])/sum(misc[g,]) * prior[g]

# 2)
LdaCV.data = lda(data, labels, CV=TRUE) # specify the argument CV
table(class.true=labels, class.assignedCV=LdaCV.data$class)
errorsCV = (LdaCV.data$class != labels)
AERCV = sum(errorsCV)/length(labels)
AERCV
```

### **2.3.** Plot the partition induced by LDA
```{r, warning=FALSE, message=FALSE}
plot(data, main='Iris Sepal', xlab='X1', ylab='X1', pch=20)
points(data[i1,], col='red', pch=20)
points(data[i2,], col='green', pch=20)
points(data[i3,], col='blue', pch=20)
legend("topright", legend=levels(labels), fill=c('red','green','blue'), cex=.7)
points(lda.data$means, pch=4,col=c('red','green','blue') , lwd=2, cex=1.5)

x  = seq(min(data[,1]), max(iris[,1]), length=200)
y  = seq(min(data[,2]), max(iris[,2]), length=200)
xy = expand.grid(X1=x, X2=y)
z  = predict(lda.data, xy)$post # these are P_i*f_i(x,y)
z1 = z[,1] - pmax(z[,2], z[,3]) # P_1*f_1(x,y)-max{P_j*f_j(x,y)}
z2 = z[,2] - pmax(z[,1], z[,3]) # P_2*f_2(x,y)-max{P_j*f_j(x,y)}
z3 = z[,3] - pmax(z[,1], z[,2]) # P_3*f_3(x,y)-max{P_j*f_j(x,y)}

# Plot the contour line of level (levels=0) of z1, z2, z3:
# P_i*f_i(x,y)-max{P_j*f_j(x,y)}=0 i.e., boundary between R.i and R.j
# where j realizes the max.
contour(x, y, matrix(z1, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z2, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z3, 200), levels=0, drawlabels=F, add=T)
```

### **2.4.** 3D plot of the partition induced by LDA
```{r, warning=FALSE, message=FALSE}
# TRY IT USING SCRIPT !
#open3d()
#points3d(data[i1,1], data[i1,2], 0, col='red', pch=15)
#points3d(data[i2,1], data[i3,2], 0, col='green', pch=15)
#points3d(data[i3,1], data[i2,2], 0, col='blue', pch=15)
#surface3d(x,y,matrix(dmvnorm(xy, m1, Sp) / 3, 50), alpha=0.4, color='red')
#surface3d(x,y,matrix(dmvnorm(xy, m2, Sp) / 3, 50), alpha=0.4, color='green', add=T)
#surface3d(x,y,matrix(dmvnorm(xy, m3, Sp) / 3, 50), alpha=0.4, color='blue', add=T)
#box3d()
```

### **2.5.** QDA (multivariate)
```{r, warning=FALSE, message=FALSE}
qda.data = qda(data, labels)
qda.data
Qda.data = predict(qda.data, data)

# APER
# Remark: correct only if we estimate the priors through the sample frequencies!
table(class.true=labels, class.assigned=Qda.data$class)
errorsq = (Qda.data$class != labels)
APERq = sum(errorsq)/length(labels)

# AER
QdaCV.data = qda(data, labels, CV=T)
errorsqCV = (QdaCV.data$class != labels)
AERqCV = sum(errorsqCV)/length(labels)
```

### **2.6.** Plot the partition induced by QDA
```{r, warning=FALSE, message=FALSE}
plot(data, main='Main Title', xlab='X1', ylab='X2', pch=20)
points(data[i1,], col='red', pch=20)
points(data[i2,], col='green', pch=20)
points(data[i3,], col='blue', pch=20)
legend("topright", legend=levels(labels), fill=c('red','green','blue'))
points(qda.data$means, col=c('red','green','blue'), pch=4, lwd=2, cex=1.5)

x  = seq(min(iris[,1]), max(data[,1]), length=200)
y  = seq(min(iris[,2]), max(data[,2]), length=200)
xy = expand.grid(X1=x, X2=y)
z  = predict(qda.data, xy)$post
z1 = z[,1] - pmax(z[,2], z[,3])
z2 = z[,2] - pmax(z[,1], z[,3])
z3 = z[,3] - pmax(z[,1], z[,2])

contour(x, y, matrix(z1, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z2, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z3, 200), levels=0, drawlabels=F, add=T)
```

### **2.7.** 3D plot of the partition induced by LDA
```{r, warning=FALSE, message=FALSE}
# TRY IT USING SCRIPT !
#open3d()
#points3d(data[i1,1], data[i1,2], 0, col='red', pch=15)
#points3d(data[i2,1], data[i3,2], 0, col='green', pch=15)
#points3d(data[i3,1], data[i2,2], 0, col='blue', pch=15)
#surface3d(x,y,matrix(dmvnorm(xy, m1, S1) / 3, 50), alpha=0.4, color='red')
#surface3d(x,y,matrix(dmvnorm(xy, m2, S2) / 3, 50), alpha=0.4, color='green', add=T)
#surface3d(x,y,matrix(dmvnorm(xy, m3, S3) / 3, 50), alpha=0.4, color='blue', add=T)
#box3d()
```

### **2.8.** KNN
```{r, warning=FALSE, message=FALSE}
# Plot the partition induced by knn
k = 7
plot(data, main='Iris.Sepal', xlab='X1', ylab='X2', pch=20)
points(data[i1,], col=2, pch=20)
points(data[i3,], col=4, pch=20)
points(data[i2,], col=3, pch=20)
legend("topright", legend=levels(labels), fill=c(2,3,4))

x  = seq(min(data[,1]), max(data[,1]), length=200)
y  = seq(min(data[,2]), max(data[,2]), length=200)
xy = expand.grid(Sepal.Length=x, Sepal.Width=y)
data.knn = knn(train = data, test = xy, cl = labels, k = k)
z = as.numeric(data.knn)
contour(x, y, matrix(z, 200), levels=c(1.5, 2.5), drawlabels=F, add=T)
```

### **2.9.** Fisher's argument 
```{r, warning=FALSE, message=FALSE}
# Let's look at the directions that highlight the discrimination among groups:
# => canonical directions

# covariance between groups (estimate)
B = 1/g*(cbind(m1 - m) %*% rbind(m1 - m) +
         cbind(m2 - m) %*% rbind(m2 - m) +
         cbind(m3 - m) %*% rbind(m3 - m))
B

# Covariance within groups (estimate)
Sp
s = min(g-1,p)
val.Sp = eigen(Sp)$val
vec.Sp = eigen(Sp)$vec
invSp.2 = 1/sqrt(val.Sp[1])*vec.Sp[,1]%*%t(vec.Sp[,1]) +
  1/sqrt(val.Sp[2])*vec.Sp[,2]%*%t(vec.Sp[,2])
invSp.2

# Spectral decomposition of Sp^(-1/2) B Sp^(-1/2)
spec.dec = eigen(invSp.2 %*% B %*% invSp.2)

# First canonical coordinate
a1 = invSp.2 %*% spec.dec$vec[,1]
a1

# second canonical coordinate
a2 = invSp.2 %*% spec.dec$vec[,2]
a2

# compare with the output of lda():
lda.data
```

#### **2.9.1.** How are the data classified?
```{r, warning=FALSE, message=FALSE}
# Compute the canonical coordinates of the data
cc1.data = as.matrix(data)%*%a1
cc2.data = as.matrix(data)%*%a2
coord.cc = cbind(cc1.data,cc2.data)

# Compute the coordinates of the mean within groups along the canonical directions
cc.m1 = c(m1%*%a1, m1%*%a2)
cc.m2 = c(m2%*%a1, m2%*%a2)
cc.m3 = c(m3%*%a1, m3%*%a2)

# Assign data to groups
f.class=rep(0, n)
for(i in 1:n) # for each datum
{
  # Compute the Euclidean distance of the i-th datum from mean within the groups
  dist.m=c(d1=sqrt(sum((coord.cc[i,]-cc.m1)^2)),
           d2=sqrt(sum((coord.cc[i,]-cc.m2)^2)),
           d3=sqrt(sum((coord.cc[i,]-cc.m3)^2)))
  # Assign the datum to the group whose mean is the nearest
  f.class[i]=which.min(dist.m)
}
f.class

table(class.true=labels, class.assigned=f.class)
errors = n - sum(diag(table(class.true=labels, class.assigned=f.class)))
APERf = errors/length(labels)
```

#### **2.9.2.** How do we classify a new observation?
```{r, warning=FALSE, message=FALSE}
x.new = c(5.85, 2.90)

# Compute the canonical coordinates
cc.new = c(x.new%*%a1, x.new%*%a2)

# Compute the distance from the means
dist.m = c(d1=sqrt(sum((cc.new-cc.m1)^2)),
           d2=sqrt(sum((cc.new-cc.m2)^2)),
           d3=sqrt(sum((cc.new-cc.m3)^2)))

# Assign to the nearest mean
which.min(dist.m)
```

#### **2.9.3.** Visually
```{r, warning=FALSE, message=FALSE}
color.species=rep(c('red','green','blue'), each=50)
iris2 = data
species.name = labels
par(mfrow=c(1,2))
plot(iris2[,1], iris2[,2], main='Plane of original coordinates',
     xlab='Sepal.Length', ylab='Sepal.Width', pch=20, col=as.character(color.species))
legend("topleft", legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
points(x.new[1], x.new[2], col='gold', pch=19)
points(m1[1], m1[2], pch=4,col='red' , lwd=2, cex=1.5)
points(m2[1], m2[2], pch=4,col='green' , lwd=2, cex=1.5)
points(m3[1], m3[2], pch=4,col='blue' , lwd=2, cex=1.5)
plot(cc1.data, cc2.data, main='Plane of canonical coordinates',
     xlab='first canonical coordinate', ylab='second canonical coordinate',
     pch=20, col=as.character(color.species))
legend("topleft", legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
points(cc.m1[1], cc.m1[2], pch=4,col='red' , lwd=2, cex=1.5)
points(cc.m2[1], cc.m2[2], pch=4,col='green' , lwd=2, cex=1.5)
points(cc.m3[1], cc.m3[2], pch=4,col='blue' , lwd=2, cex=1.5)
points(cc.new[1], cc.new[2], col='gold', pch=19)
segments(cc.m1[1], cc.m1[2], cc.new[1], cc.new[2])
segments(cc.m2[1], cc.m2[2], cc.new[1], cc.new[2])
segments(cc.m3[1], cc.m3[2], cc.new[1], cc.new[2])
```

#### **2.9.4.** Plot the partition generated by the canonical coordinates
```{r, warning=FALSE, message=FALSE}
cc1.iris = cc1.data
cc2.iris = cc2.data
color.species = species.name
levels(color.species) = c('red','green','blue')
plot(cc1.iris, cc2.iris, main='Fisher discriminant analysis',
     xlab='first canonical coordinate', ylab='second canonical coordinate',
     pch=20, col=as.character(color.species))
legend("topleft", legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
points(cc.m1[1], cc.m1[2], pch=4,col='red' , lwd=2, cex=1.5)
points(cc.m2[1], cc.m2[2], pch=4,col='green' , lwd=2, cex=1.5)
points(cc.m3[1], cc.m3[2], pch=4,col='blue' , lwd=2, cex=1.5)
x.cc  = seq(min(cc1.iris),max(cc1.iris),len=200)
y.cc  = seq(min(cc2.iris),max(cc2.iris),len=200)
xy.cc = expand.grid(cc1=x.cc, cc2=y.cc)
z     = cbind( sqrt(rowSums(scale(xy.cc,cc.m1,scale=FALSE)^2)),
               sqrt(rowSums(scale(xy.cc,cc.m2,scale=FALSE)^2)),
               sqrt(rowSums(scale(xy.cc,cc.m3,scale=FALSE)^2)))
z1.cc = z[,1] - pmin(z[,2], z[,3])
z2.cc = z[,2] - pmin(z[,1], z[,3])
z3.cc = z[,3] - pmin(z[,1], z[,2])
contour(x.cc, y.cc, matrix(z1.cc, 200), levels=0, drawlabels=F, add=T)
contour(x.cc, y.cc, matrix(z2.cc, 200), levels=0, drawlabels=F, add=T)
contour(x.cc, y.cc, matrix(z3.cc, 200), levels=0, drawlabels=F, add=T)
```

#### **2.9.5** Plot LDA
```{r, warning=FALSE, message=FALSE}
lda.iris = lda.data
plot(iris2, main='Iris Sepal', xlab='Sepal.Length', ylab='Sepal.Width', pch=20)
points(iris2[i1,], col='red', pch=20)
points(iris2[i2,], col='green', pch=20)
points(iris2[i3,], col='blue', pch=20)
legend("topright", legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
points(rbind(m1,m2,m3), pch=4,col=c('red','green','blue') , lwd=2, cex=1.5)

x  = seq(min(iris[,1]), max(iris[,1]), length=200)
y  = seq(min(iris[,2]), max(iris[,2]), length=200)
xy = expand.grid(Sepal.Length=x, Sepal.Width=y)
z  = predict(lda.iris, xy)$post # these are P_i*f_i(x,y)
z1 = z[,1] - pmax(z[,2], z[,3]) # P_1*f_1(x,y)-max{P_j*f_j(x,y)}
z2 = z[,2] - pmax(z[,1], z[,3]) # P_2*f_2(x,y)-max{P_j*f_j(x,y)}
z3 = z[,3] - pmax(z[,1], z[,2]) # P_3*f_3(x,y)-max{P_j*f_j(x,y)}

# Plot the contour line of level (levels=0) of z1, z2, z3:
# P_i*f_i(x,y)-max{P_j*f_j(x,y)}=0 i.e., boundary between R.i and R.j
# where j realizes the max.
contour(x, y, matrix(z1, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z2, 200), levels=0, drawlabels=F, add=T)
contour(x, y, matrix(z3, 200), levels=0, drawlabels=F, add=T)
```

#### **2.9.6.** Plot of the projections on the canonical directions (not orthogonal!)
```{r, warning=FALSE, message=FALSE}
plot(iris2, main='Projection on the canonical directions',
     xlab='Sepal.Length', ylab='Sepal.Width', pch=20, xlim=c(-3,8), ylim=c(-3,7))
points(iris2[i1,], col='red', pch=20)
points(iris2[i2,], col='green', pch=20)
points(iris2[i3,], col='blue', pch=20)
legend('topleft', legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
points(rbind(m1,m2,m3), pch=4,col=c('red','green','blue') , lwd=2, cex=1.5)
abline(h=0,v=0, col='grey35')
arrows(x0=0, y0=0, x1=a1[1], y1=a1[2], length=.1)
arrows(x0=0, y0=0, x1=a2[1], y1=a2[2], length=.1)
text(a1[1], a1[2], 'a1',pos=3)
text(a2[1], a2[2], 'a2',pos=2)
abline(coef=c(0,(a1[2]/a1[1])), col='grey55',lty=2)
abline(coef=c(0,(a2[2]/a2[1])), col='grey55',lty=2)
points(cc1.iris*a1[1]/(sum(a1^2)),cc1.iris*a1[2]/(sum(a1^2)),
       col=as.character(color.species))
points(cc2.iris*a2[1]/(sum(a2^2)),cc2.iris*a2[2]/(sum(a2^2)),
       col=as.character(color.species))
plot(cc1.iris, cc2.iris, main='Coordinate system of the canonical coordinates',
     xlab='first canonical coordinate', ylab='second canonical coordinate',
     pch=20, col=as.character(color.species))
legend('topleft', legend=levels(species.name), fill=c('red','green','blue'), cex=.7)
cc.m1 = c(m1%*%a1, m1%*%a2)
cc.m2 = c(m2%*%a1, m2%*%a2)
cc.m3 = c(m3%*%a1, m3%*%a2)
points(cc.m1[1], cc.m1[2], pch=4,col='red' , lwd=2, cex=1.5)
points(cc.m2[1], cc.m2[2], pch=4,col='green' , lwd=2, cex=1.5)
points(cc.m3[1], cc.m3[2], pch=4,col='blue' , lwd=2, cex=1.5)
```

